---
title: "Comparison of clustering results"
author: "Ying Jiang and Joel Eliason"
date: "2021/11/15"
output: html_document
---

The 10x_spatial_positions and X_pca files are the files of the spatial positions of the cells and the reduced-dim features for each cell.

df$ncluster: 3/6/10
df$lambda: 1/10
df$quantile_fun: colMaxs/colMeds/colMins 

```{r setup, include=FALSE}
X = read.csv("X_pca.csv", header = F) # 50-d features of cells after PCA
coord = read.csv("10x_spatial_positions.csv", header = F)# 2-dimensional position
df = readRDS("cluster_sweep_results.rds") # results from running the cluster_sweep function

# change 'cluster' variable from factor to numerical
df$cluster =as.numeric(levels(df$cluster)[df$cluster])
```

## Comparision between Different clustering methods: 
1. basic KMeans 
2. hierarchical 
3. GMM

All methods are on X(50-d after PCA) and use coord to do visualization
```{r}
library(stats)
library(flexclust)
library(mclust)
library(cowplot)
library(cluster)
library(knitr)
library(tidyverse)
```

#### A func to get clustering results from different clustering methods.
```{r}
#input:
## k: 3/6/10
## l: 1/10
## q_fun: 'colMaxs'/'colMeds'/'colMins' 

#output:
## dataframe: cols: x - coord x
##                  y - coord y
##                  kmeans_spatial - kmeans_spatial clustering result with specific k/lambda/quantile function
##                  kmeans - kmeans clustering result
##                  hier - hierarchical clustering result
##                  gmm - GMM clustering result

compare = function(k, l, q_fun){
  data = df %>% filter(lambda == l) %>%
         filter(nclust == k) %>%
         filter(quantile_fun == q_fun)
  
  # basic k-means
  kmeans_res = kmeans(X, k)$cluster
  
  #  hierarchical
  X.scale = scale(X)
  d = dist(X.scale)
  fit<-hclust(d, method="ward.D2") # Ward method
  hier_res = cutree(fit, k)
  
  # GMM
  gmm_res = Mclust(X, G=k)$classification
  
  data_show = cbind(coord, data$cluster, kmeans_res, hier_res, gmm_res)
  colnames(data_show) = c('x', 'y', 'kmeans_spatial', 'kmeans', 'hier', 'gmm')
  return(data_show)
}

compare_precomp = function(k, l, q_fun, precomp_data){
  data = df %>% filter(lambda == l) %>%
         filter(nclust == k) %>%
         filter(quantile_fun == q_fun)
  
  data_show = cbind(coord, data$cluster, precomp_data)
  colnames(data_show) = c('x', 'y', 'kmeans_spatial', 'kmeans', 'hier', 'gmm')
  return(data_show)
}
```

```{r}
# k=3/6/10, lambda=1, quantile_function = 'colMeds' -- Can also try other parameters
#will take some time due to the GMM model
data.3 = compare(3, 1, 'colMeds')
data.6 = compare(6, 1, 'colMeds')
data.10 = compare(10, 1, 'colMeds')
```

```{r}
# if data are already computed from above, can reload
# data = readRDS("comparisons.rds")
# data.3 = data[[1]]
# data.6 = data[[2]]
# data.10 = data[[3]]
# precomp.3 = data.3[,4:6]
# precomp.6 = data.6[,4:6]
# precomp.10 = data.10[,4:6]
```


```{r}
# plot k=3 as an example
p1 = ggplot(data.3) + geom_point(aes(x = x, y = y, colour = factor(kmeans_spatial)))+ggtitle('Kmeans spatial') + theme(legend.position = "none")
p2 = ggplot(data.3) + geom_point(aes(x = x, y = y, colour = factor(kmeans)))+ggtitle('Kmeans') + theme(legend.position = "none")
p3 = ggplot(data.3) + geom_point(aes(x = x, y = y, colour = factor(hier)))+ggtitle('Hierarchical') + theme(legend.position = "none")
p4 = ggplot(data.3) + geom_point(aes(x = x, y = y, colour = factor(gmm)))+ggtitle('GMM') + theme(legend.position = "none")

# p1;p2;p3;p4

p = plot_grid(p1,p2,p3,p4)
save_plot('methods_spatial_plot.png',p)
```
The result of kmeans_spatial, kmeans and gmm are similar, pretty different from result of hierarchical clustering.

Things to note:
* The result of kmeans_spatial, kmeans and gmm are pretty similar, pretty different from result of hierarchical clustering.
* Can see that k-means spatial makes groups much more cohesive

Let's also look at k-means spatial for three clusters for other parameterizations:

```{r}
df %>% 
  filter(nclust == 3) %>%
  filter(lambda == 1) %>%
  mutate(cluster = factor(cluster)) %>%
  ggplot(aes(x=x,y=y)) + 
  geom_point(aes(colour = cluster)) +
  facet_wrap(.~ quantile_fun) +
  theme(aspect.ratio=1)

ggsave('spatial_by_quantile.pdf')

system2(command = "pdfcrop", 
        args    = c("spatial_by_quantile.pdf", 
                    "spatial_by_quantile_crop.pdf") 
        )
```

Things to note:

* Between lambda values, colMins: Clusters are more spotted for large lambda (lambda = 10). May be because colMins is too small.
* colMeds: Again, looks like clusters are more spotted for lambda = 10. Also big section in the middle is split (could be an initialization problem)
* colMaxs: Most notable is the presence of basically two clusters in lambda = 10. Could again be an initialization problem. That little peach piece at the bottom doesn't seem to show up either in lambda = 1.
* It seems with lambda = 10 that more spot-formation is happening since the lambda is pushing even spatially close 

colMins seems to be quite close to normal k-means, in terms of spotting. However, can already see some cohesion happening. This cohesion seems to only increase across the other 2 quantile functions. Larger and larger patches of undisturbed clusters.

Things to improve
* speed
* initialization (and also consistency of clustering - aka want to preserve self-ARI)
* Run all of these multiple times each to be able to see consistency of ARI


#### Metrics through different methods.
1. adjusted rand index

3x3 matrix:
similarity between kmeans_spatial method and other methods
row: k=3/6/10
col: kmeans/hier/gmm
```{r}
data = list(data.3, data.6, data.10)
ad_rand_ind = matrix(0, nrow = 3, ncol=3)

for(i in 1:3){
  for(j in 1:3){
    ad_rand_ind[i,j] = adjustedRandIndex(data[[i]]$kmeans_spatial, data[[i]][, (3+j)])
  }
}
colnames(ad_rand_ind) = c('kmeans', 'hier', 'gmm')
rownames(ad_rand_ind) = c('k=3', 'k=6', 'k=10')
ad_rand_ind

```

```{r}
ad_rand_ind %>%
  as_tibble(rownames = "k") %>%
  pivot_longer(cols = kmeans:gmm,names_to = "method",values_to = "ARI") %>%
  mutate(k = as.numeric(substring(k,3))) %>%
  mutate(method = factor(method)) %>%
  ggplot() +
    geom_line(aes(k,ARI,colour = method,group = method)) +
    labs(title = "ARI of spatial k-means (colMeds, lambda = 1) with basic clustering methods")
  
```

```{r}
k.list = c(3, 6, 10)
qfunc.list = c('colMaxs','colMeds','colMins')

grid = expand.grid(qfunc.list,k.list)
aris = apply(grid,1,function(g) {
  sapply(dfs, function(d) {
  x = d %>%
    filter(nclust == as.numeric(g[[2]])) %>%
    filter(quantile_fun == g[[1]])
    list(ari_kmeans=adjustedRandIndex(x$cluster,x$cluster_kmeans),ari_hier=adjustedRandIndex(x$cluster,x$cluster_hier))
})
})

mean_aris = sapply(aris,function(ari) {
  list(ari_kmeans=mean(unlist(ari[1,])),ari_hier=mean(unlist(ari[2,])))
})

sd_aris = sapply(aris,function(ari) {
  list(ari_kmeans=sd(unlist(ari[1,])),ari_hier=sd(unlist(ari[2,])))
})

ari.df = cbind(grid,unlist(mean_aris[1,]),unlist(mean_aris[2,]),unlist(sd_aris[1,]),unlist(sd_aris[2,]))

colnames(ari.df) = c('quantile_fun','nclust','ARI_kmeans','ARI_hier','sd_kmeans','sd_hier')

ari.df %>%
  as_tibble() %>%
  pivot_longer(cols = -c("quantile_fun","nclust"), names_to = c(".value","method"),names_sep = "_") %>%
  mutate(method = factor(method)) %>%
  ggplot(aes(x = nclust,colour = quantile_fun)) +
    geom_line(aes(y = ARI)) + 
    geom_errorbar(aes(ymin=ARI-sd,ymax=ARI+sd,color=quantile_fun,width=0.2)) +
    facet_grid(.~method)

ggsave('sim_ARI.png')
```

2. silhouette

3x4 matrix:
For each model, compute the total mean of individual silhouette widths
row: k=3/6/10
col: kmeans_spatial/kmeans/hier/gmm
```{r}
dis = dist(X)^2
si = matrix(0, nrow = 3, ncol=4)
for(i in 1:3){
  for(j in 1:4){
    si_res = silhouette(data[[i]][,(2+j)], dis)
    si[i, j] = summary(si_res)$avg.width # use the total mean of individual silhouette widths to compare models
  }
}
colnames(si) = c('kmeans_spatial','kmeans', 'hier', 'gmm')
rownames(si) = c('k=3', 'k=6', 'k=10')
si
```

```{r}
si %>%
  as_tibble(rownames = "k") %>%
  pivot_longer(cols = kmeans_spatial:gmm,names_to = "method",values_to = "Silhouette") %>%
  mutate(k = as.numeric(substring(k,3))) %>%
  mutate(method = factor(method)) %>%
  ggplot() +
    geom_line(aes(k,Silhouette,colour = method,group = method)) +
    labs(title = "Average silhouette width of various methods")

ggsave('avg_sil.png')
```





The performance of kmeans_spatial is not as good as basic kmeans, while it is better than hierarchical and gmm.

3. Within cluster sum of squares
```{r}
# function of computing sum of squares in clusters
# input: k, result - clustering result
css = function(k, result){
  total = 0
  for(i in 1:k){
    x = X[result==i, ]
    n = nrow(x)
    x.mean = colMeans(x)
    x = x-matrix(rep(x.mean, n), byrow = T, nrow=n)
    total = total + sum(x^2)
  }
  return(total)
}

k.list = c(3, 6, 10)
sum.sq = matrix(0, nrow = 3, ncol=4)
for(i in 1:3){
  for(j in 1:4){
    sum.sq[i, j] = css(k.list[i], data[[i]][, 2+j]) 
  }
}
colnames(sum.sq) = c('kmeans_spatial','kmeans', 'hier', 'gmm')
rownames(sum.sq) = c('k=3', 'k=6', 'k=10')
sum.sq
```


```{r}
sum.sq %>%
  as_tibble(rownames = "k") %>%
  pivot_longer(cols = kmeans_spatial:gmm,names_to = "method",values_to = "WCSS") %>%
  mutate(k = as.numeric(substring(k,3))) %>%
  mutate(method = factor(method)) %>%
  ggplot() +
    geom_line(aes(k,WCSS,colour = method,group = method)) +
    labs(title = "WCSS of various clustering methods")
```

4. Avg distance from centroid

```{r}
# function of computing sum of squares in clusters
# input: k, result - clustering result
avg_css = function(k, result){
  total = 0
  for(i in 1:k){
    x = X[result==i, ]
    n = nrow(x)
    x.mean = colMeans(x)
    x = x-matrix(rep(x.mean, n), byrow = T, nrow=n)
    total = total + mean(sqrt(rowSums(x^2)))
  }
  return(total/k)
}

k.list = c(3, 6, 10)
avg.sq = matrix(0, nrow = 3, ncol=4)
for(i in 1:3){
  for(j in 1:4){
    avg.sq[i, j] = avg_css(k.list[i], data[[i]][, 2+j]) 
  }
}

colnames(avg.sq) = c('kmeans_spatial','kmeans', 'hier', 'gmm')
rownames(avg.sq) = c('k=3', 'k=6', 'k=10')
avg.sq
```

```{r}
avg.sq %>%
  as_tibble(rownames = "k") %>%
  pivot_longer(cols = kmeans_spatial:gmm,names_to = "method",values_to = "Avg_SS") %>%
  mutate(k = as.numeric(substring(k,3))) %>%
  mutate(method = factor(method)) %>%
  ggplot() +
    geom_line(aes(k,Avg_SS,colour = method,group = method)) +
    labs(title = "Average SS of various clustering methods")

ggsave('avg_ss.png')
```


## Comparision between different nclust/lambda/quantile_func of kmeans_spatial methods
1. silhouette
```{r}
si.list  = list()

k.list = c(3, 6, 10)
lambda.list = c(1, 10)
qfunc.list = c('colMaxs','colMeds','colMins')

for(k in 1:3){
  si_k = matrix(0, nrow= 2, ncol=3)
  colnames(si_k) = qfunc.list
  rownames(si_k) = c("lambda=1", "lambda=10")
  
for(i in 1:2){
  for(j in 1:3){
    temp =  df %>% filter(lambda == lambda.list[i]) %>%
         filter(nclust == k.list[k]) %>%
         filter(quantile_fun == qfunc.list[j])
    s = silhouette(temp$cluster, dis)
    si_k[i,j] = summary(s)$avg.width
  }
  si.list[[k]] = data.frame(si_k)
}
}
names(si.list) = c("k=3", "k=6", "k=10")
si.list
```

```{r}
si.df = do.call(rbind,si.list) %>%
  as_tibble(rownames = "paras") %>%
  separate(paras,into = c("k","lambda"),sep = "\\.") %>%
  mutate(k = as.numeric(substring(k,3))) %>%
  mutate(lambda = as.numeric(substring(lambda,8))) %>%
  pivot_longer(cols = colMaxs:colMins,names_to = "quant_fun",values_to = "Silhouette") %>%
  mutate(quant_fun = factor(quant_fun)) %>%
  mutate(lambda = factor(lambda))

si.df %>%
  ggplot() +
    geom_line(aes(k,Silhouette, colour = quant_fun,linetype = lambda)) +
    labs(title = "Avg silhouette width for various parameterizations")

ggsave('sil_width_params.png')
```

2.Within cluster sum of squares
```{r}
sum.sq.list  = list()

k.list = c(3, 6, 10)
lambda.list = c(1, 10)
qfunc.list = c('colMaxs','colMeds','colMins')
for(k in 1:3){
  sum.sq_k = matrix(0, nrow= 2, ncol=3)
  colnames(sum.sq_k) = qfunc.list
  rownames(sum.sq_k) = c("lambda=1", "lambda=10")
  
for(i in 1:2){
  for(j in 1:3){
    temp =  df %>% filter(lambda == lambda.list[i]) %>%
         filter(nclust == k.list[k]) %>%
         filter(quantile_fun == qfunc.list[j])
    sum.sq_k[i,j] = css(k.list[k], temp$cluster)
  }
  sum.sq.list[[k]] = data.frame(sum.sq_k)
}
}
names(sum.sq.list) = c("k=3", "k=6", "k=10")
sum.sq.list
```


```{r}
sum.sq.df = do.call(rbind,sum.sq.list) %>%
  as_tibble(rownames = "paras") %>%
  separate(paras,into = c("k","lambda"),sep = "\\.") %>%
  mutate(k = as.numeric(substring(k,3))) %>%
  mutate(lambda = as.numeric(substring(lambda,8))) %>%
  pivot_longer(cols = colMaxs:colMins,names_to = "quant_fun",values_to = "WCSS") %>%
  mutate(quant_fun = factor(quant_fun)) %>%
  mutate(lambda = factor(lambda))

sum.sq.df %>%
  ggplot() +
    geom_line(aes(k,WCSS, colour = quant_fun,linetype = lambda)) +
    labs(title = "Avg WCSS for various parameterizations")
```

3. Average sum of squares
```{r}
avg.sq.list  = list()

k.list = c(3, 6, 10)
lambda.list = c(1, 10)
qfunc.list = c('colMaxs','colMeds','colMins')
for(k in 1:3){
  sum.sq_k = matrix(0, nrow= 2, ncol=3)
  colnames(sum.sq_k) = qfunc.list
  rownames(sum.sq_k) = c("lambda=1", "lambda=10")
  
for(i in 1:2){
  for(j in 1:3){
    temp =  df %>% filter(lambda == lambda.list[i]) %>%
         filter(nclust == k.list[k]) %>%
         filter(quantile_fun == qfunc.list[j])
    sum.sq_k[i,j] = avg_css(k.list[k], temp$cluster)
  }
  avg.sq.list[[k]] = data.frame(sum.sq_k)
}
}
names(avg.sq.list) = c("k=3", "k=6", "k=10")
avg.sq.list
```

```{r}
avg.sq.df = do.call(rbind,avg.sq.list) %>%
  as_tibble(rownames = "paras") %>%
  separate(paras,into = c("k","lambda"),sep = "\\.") %>%
  mutate(k = as.numeric(substring(k,3))) %>%
  mutate(lambda = as.numeric(substring(lambda,8))) %>%
  pivot_longer(cols = colMaxs:colMins,names_to = "quant_fun",values_to = "Avg_SS") %>%
  mutate(quant_fun = factor(quant_fun)) %>%
  mutate(lambda = factor(lambda))

avg.sq.df %>%
  ggplot() +
    geom_line(aes(k,Avg_SS, colour = quant_fun,linetype = lambda)) +
    labs(title = "Avg SS for various parameterizations")
ggsave('avg_ss_params.png')
```

```{r}
saveRDS(data,"comparisons.rds")
```

